<!--- Timestamp to trigger book rebuilds: `r Sys.time()` --->

```{r setup, include = FALSE}
library(tidyverse)
# fit <- targets::tar_read(fit)
# fit
```

## May 2023

<small>Source: <code>`r knitr::current_input()`</code></small>


### Nonlinear model zoo

#### Gompertz 

The [Gompertz function](https://en.wikipedia.org/wiki/Gompertz_function)
"describes growth as being slowest at the start and end of a given time
period. The right-side or future value asymptote of the function is
approached much more gradually by the curve than the left-side or lower
valued asymptote." 

$$
f(t) = a\mathrm{e}^{-b\mathrm{e}^{-ct}} \\
a: \textrm{asymptote} \\
b: \textrm{translation left or right} \\
c: \textrm{growth factor} \\
$$

So we have two different growth rates but a single parameter relating to
growth rate. R's implementation (`stats::SSgompertz()`) "simplifies" the
double exponential into `Asym * exp(-b2 * b3 ^ x)`.

The scaling factor c or b3 makes sense in the original Gompertz
equation. I couldn't understand what the R version was doing because it was
changing the base of exponentiation.

```{r gomp0}
library(tidyverse)

gompertz <- function(xs, asym, b2, b3) {
  asym * exp(-b2 * exp(-b3 * xs))
}
stat_gompertz <- function(focus, asym, b2, b3) {
  args <- list(asym = asym, b2 = b2, b3 = b3)
  color <- sprintf("%s = %s", focus, args[[focus]])
  stat_function(
    aes(color = color),
    fun = gompertz, 
    args = args
  )
}

ggplot() + 
  xlim(-5, 5) +
  stat_gompertz("b3", 1, 1, 1) +
  stat_gompertz("b3", 1, 1, 0) +
  stat_gompertz("b3", 1, 1, .1) + 
  stat_gompertz("b3", 1, 1, -.5)
```

We have this weird thing where $b = 1/2$ and $b = 2$ are the same
distance from $b = 1$.

```{r gomp1}
ggplot() + 
  xlim(-5, 5) +
  stat_gompertz("b2", 1, 1/2, 1) +
  stat_gompertz("b2", 1, 2, 1) +
  stat_gompertz("b2", 1, 1, 1) + 
  stat_gompertz("b2", 1, 0, -1) 
```


```{r, include = FALSE, eval = FALSE}
# DNase
ggplot(DNase) + 
  aes(x = log(conc), y = density) + 
  geom_point() +
  geom_line(aes(group = Run))

library(lme4)

# Demo

DNase$log_conc <- DNase$conc |> log()

starts <- getInitial(
  density ~ SSgompertz(log_conc, Asym, b2, b3), 
  data = DNase
)

nm1 <- nlmer(
  density ~ SSgompertz(log_conc, Asym, b2, b3) ~ (b3 + Asym | Run),
  data = DNase, 
  start = starts
)

coef(nm1)[["Run"]] |> 
  tibble::rownames_to_column("Run")

DNase |> 
  tidy
# tibble <- 
#   
#   tidyr::expand(londseq(-3.5, 4, by = .1)


```

