<!--- Timestamp to trigger book rebuilds: `r Sys.time()` --->

```{r setup, include = FALSE}
# library(tidyverse)
# fit <- targets::tar_read(fit)
# fit
```

## January 2024

<small>Source: <code>`r knitr::current_input()`</code></small>


### MFA note: don't touch the solver

My colleague and I independently ran into an issue where the Montreal
Forced Aligner installation with conda seemed to take forever during the
solving environment stage. We both tried to fix this problem by changing
the default solver. This change did lead to improved installation, but
the MFA could not run at all. The correct thing to do is **do nothing**.
Just use the default solver and wait it out.

### targets + crew

In targets, by default, targets are dispatched to be built one at a
time, but we can use crew so that multiple workers can build targets in
parallel. Here is the code to add to `_targets.R` to get 2 workers.

```{r eval = FALSE}
targets::tar_option_set(
  controller = crew::crew_controller_local(workers = 2)
)
```

And then you can view how much work each worker did. (I'm hard-coding
the returns to show what happened with a full sit build.)

```{r, eval = FALSE}
targets::tar_crew()
#>   controller               worker seconds targets
#>   <chr>                     <int>   <dbl>   <int>
#> 1 8f28e966dd6adaf60895eb01      1    49.4      32
#> 2 8f28e966dd6adaf60895eb01      2    25.1       9
```

### revelation while reading withr 3.0.0 release

I was reading the [withr 3.0.0 release
notes](https://www.tidyverse.org/blog/2024/01/withr-3-0-0/) and noticed
a pleasing symmetry the `with_` and `local_` functions.

The `local_` functions set a temporary value within some local scope,
like a `local()` or a `function()`.

```{r}
local({
  withr::local_language("fr")
  plot[[1]]
})
plot[[1]]
```

The `with_` functions set a temporary value around some block of code.

```{r}
withr::with_language("fr", {
  plot[[1]]
})
plot[[1]]
```

So where are we working from?

- local: within the code (scope)
- withr: around the code

### a silly function

I got sick of writing null-guard clauses like:

```{r}
my_cool_function <- function(data, ...) {
  if (is.null(data)) { 
    return(NULL)
  }
  
  #  rest of the function
  nrow(data)
}
```

So why can't I just make that a higher order function:

```{r}
nullsafe_map <- function(x, f, ...) {
  if (is.null(x)) NULL else f(x, ...)
}

mtcars |> 
  nullsafe_map(my_cool_function) |> 
  nullsafe_map(log)

NULL |> 
  nullsafe_map(my_cool_function) |> 
  nullsafe_map(log)
```

It feels like I reinvented something vaguely monad-y here, and [I posted
the
question](https://staging.bsky.app/profile/tjmahr.com/post/3kiq3hhlvis2m):
"What have I reinvented here?" Tan suggested "`purrr::map_if()` with `.p
= Negate(is.null)`?" which is cool because I had forgotten about
`purrr::map_if()`.

### the pipe placeholder works better now

[Andrew Heiss pointed out to
me](https://staging.bsky.app/profile/andrew.heiss.phd/post/3ki4b46li6l2y)
that the placeholder since 4.3.0 is a lot better. 

```{r}
quote(
  model |> _$coef |> _[1]
) 
```

Previously, I had been circumventing these functions with things like
`getElement()` or making my own pipe-friendly wrapper functions like:

```{r}
vec_element <- function(xs, i) {
  xs[[i]]
}

vec_index <- function(xs, i) {
  xs[i]
}

vec_index_into <- function(i, xs) {
  xs[i]
}

vec_replace_na <- function(xs, replacement) {
  xs[is.na(xs)] <- replacement
  xs
}

vec_remove_na <- function(xs) {
  xs[!is.na(xs)]
}

vec_set_names <- function(xs, ns) {
  names(xs) <- ns
  xs
}
```

There were two other functions alongside these helpers that weren't
meant to solve any specific piping problem. I came to adore them when
I was doing some Advent of Code puzzles for fun in base R.

```{r}
vec_which_value <- function(xs, value, negate = FALSE) {
  (xs %in% value) |>
    xor(negate) |>
    which()
}

vec_which_name <- function(xs, name, negate = FALSE) {
  vec_which_value(names(xs), name, negate)
}

letters |> 
  vec_which_value(c("a", "e", "i", "o", "u"), negate = TRUE) |> 
  vec_index_into(letters)
```

### `purrr::pluck()` is OP

From `?purrr::pluck()`:

> [`pluck()`] also accepts arbitrary accessor functions, i.e.
> functions that take an object and return some internal piece. [...]
> Compare: `accessor(x[[1]])$foo` to `pluck(x, 1, accessor, "foo")`.

